---
title: "Analysis of U.S. Storm Damage Loss"
author: "Chen Xing"
date: "3/23/2022"
output: html_document
---

## Abstract


## Data Processing
1. The R package `Tidyverse` is main tool in this analysis.
```{r, package, message=FALSE}
require(tidyverse)
```
2. Load the data download from assignment website without unzip.
```{r, load data, message=FALSE}
dt <- read_csv("repdata_data_StormData.csv.bz2")
```
3. Select and filter the data we need to find the answers to the two questions, and group it by severe weather types.
```{r, filter data}
dt_1 <- dt %>% select(BGN_DATE, BGN_TIME, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,CROPDMG, CROPDMGEXP) %>%
  group_by(EVTYPE)
head(dt_1)
```
4. Calculate the total number of fatalities and injuries between `evtype`.
```{r, fatalities and injuries}
dt_fj <- dt_1 %>% mutate(fj=FATALITIES+INJURIES) %>%
  summarise(sumfj=sum(fj), sumi=sum(INJURIES), sumf=sum(FATALITIES)) %>%
  arrange(-sumfj)
head(dt_fj)
```
5. Calculate the total economic consequences between `evtype`.
```{r, fatalities and injuries}
dt_eco <- dt_1 %>% pmap()
head(dt_fj)
```
## Results